Title

### 2018-07-31 thesis/paper/paper.tex
Modded Datalog

Other information regarding items like labels and descriptions in different languages as well as aliases and site links to other Wikimedia projects are also part of the data model but are not subject to constraints and are thus not explored further.
\section{Datalog}
Datalog is a logic programming language. It is fully declarative and uses simple rules to derive additional knowledge from given facts. The most basic unit of information in datalog is a \emph{constant}, which will be denoted as someConstant. In the context of Wikidata any item or property, e.g. Earth (Q2) or head of government (P6), could be represented as a constant. Relations in datalog are expressed as \emph{predicates} with a specified arity \(n\), written as predicate\([n]\). To stay close to Wikidata in our examples let us take the predicate statement\([3]\) as an example. This would be used to express the \emph{item}-\emph{property}-\emph{value} relation with each position representing one part of the triple.\\
Additionally, datalog also knows \emph{variables}, which will be capitalized like \(X\). Constants and variables are collectively known as \emph{terms} and are, together with predicates, used to form \emph{atoms}. An atom has the form predicate\((t_1, \ldots , t_n)\) with \(n\) as the arity of the predicate and \(t_1,\ldots,t_n\) terms.\\
Atoms are the building blocks of \emph{rules}, which are used to derive information in datalog. Given atoms \(C\) and \(B_n\), a rule takes the form \(C \leftarrow P_1 \wedge \ldots \wedge P_m\), with \(C\) being the conclusion and \(B_1 \wedge \ldots \wedge B_m\) the premise. If the premise does not contain any atoms such a rule is called a \emph{fact}, or a \emph{ground rule} if it additionally only uses constants. A finite set of rules forms a \emph{program}.\\

@@

\todo[inline]{Explain the relation to FO-logic?}

@@

\todo[inline]{Probably a good place to explain existential and universal quantifiers.}
\section{Representation}
To model the basic representation of triples, qualifiers and references three predicates are introduced.\\
tripleEDB\([4]\) represents a single statement consisting of the statement ID, the item ID, the property ID and the value as either and item ID or a string-representation of the data value.\\
qualifierEDB\([3]\) and referenceEDB\([3]\) represent the qualifiers and references on a given statement ID with their respective property id and the value as usual.\\

>>>>

Other information regarding items like labels and descriptions in different languages as well as aliases and site links to other Wikimedia projects are also part of the data model but are not subject to constraints and are thus not explored further.
\section{Datalog}
Datalog is a logic programming language. It is fully declarative and uses existential rules to derive additional knowledge from given facts. The most basic unit of information in datalog is a \emph{constant}, which will be denoted as someConstant. In the context of Wikidata any item or property, e.g. Earth (Q2) or head of government (P6), could be represented as a constant. Relations in datalog are expressed as \emph{predicates} with a specified arity \(n\), written as predicate\([n]\). To stay close to Wikidata in our examples let us take the predicate statement\([3]\) as an example. This would be used to express the \emph{item}-\emph{property}-\emph{value} relation with each position representing one part of the triple.\\
Additionally, datalog also knows \emph{variables}, which will be capitalized like \(X\). Constants and variables are collectively known as \emph{terms} and are, together with predicates, used to form \emph{atoms}. An atom has the form predicate\((t_1, \ldots , t_n)\) with \(n\) as the arity of the predicate and \(t_1,\ldots,t_n\) terms.\\
Atoms are the building blocks of \emph{existential rules}, which are used to derive information in datalog. A rule is premise that implies a conclusion, and both are conjunctions of atoms. Denote \(P_n\) as the premise atoms, \(C_n\) as the conclusion atoms. The variables used are divided into three disjunct sets \(V_p\), \(V_c\), and \(V_r\) which are the variables occurring only in premise atoms, only in conclusion atoms, and occurring in both respectively. The general from of an existential rule is (with \(k \in \mathbb{N}^+\), \(l \in \mathbb{N}\))
\begin{equation*}
\forall p \in V_p, r \in V_r : \exists c \in V_c: C_1 \ldots C_k \leftarrow P_1 \ldots P_l
\end{equation*}

\(V_c\) are called existential variables. If \(V_c = \emptyset\) the rule is called a \emph{datalog rule} and if \(l = 0\) is called a \emph{fact}. Since \(\forall p \in V_p, r \in V_r\) can be taken as given for all rules, it will be omitted in this work.

@@

@@

\todo[inline]{Should I explain the chase? And in what detail...}
\section{Representation}
To model the basic representation of triples, qualifiers and references three predicates are introduced.\\
tripleEDB\([4]\) represents a single statement consisting of the statement ID, the item ID, the property ID and the value as either and item ID or a string-representation of the data value.\\
qualifierEDB\([3]\) and referenceEDB\([3]\) represent the qualifiers and references on a given statement ID with their respective property id and the value as usual.\\
